Lake Photic Zone Temperature accross the Conterminous United States

### 2019-08-08 manuscript/manuscript_old.Rmd
trying to fix something

During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact every  aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting will become crucial for protecting human and environmental health.

Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.  These studies typically vary in number of lakes studied, complexity of modelling approach, and time interval.  Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).  Such studies include: [@saeed2016water; @peeters2002modeling; @zhong2016recent].  There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  By and large, air temperature and lake size are often the only selected predictor variables.  Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.  Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].  Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].

@@

The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).  The NLA is stratified random sample of all lakes in the conterminous United State.  The NLA was conducted in 2007 and then repeated in 2012.  Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.  The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.

@@

PARAGRAPH about testing hypotheses

@@

Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.  Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).  Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).  Model performance is reported as mean square error and adjusted R.

@@

Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.   The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`). The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.

@@

Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States.  Average ambient air temperature from the sample date, data obtained from the PrRISM Climate Group, was the most important variable imapcting the final model's accuracy.

>>>>

During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact every  aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.

Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.  These studies typically vary in number of lakes studied, complexity of modelling approach, and study period.  Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).  Such studies include: [@saeed2016water; @peeters2002modeling; @zhong2016recent].  There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  By and large, air temperature and lake size are often the only selected predictor variables.  Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.  Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].  Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].

@@

@@

@@

Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.  Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).  Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.(ADD SENTENCE)  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).  Model performance is reported as mean square error and adjusted R.

@@

Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.   The variable selection process identified a reduced model with eight variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`). The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.

@@

Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States. The final model had a mean square error of 2.19 and an adjusted R^2 of 0.88.  The sampling date and that day’s average ambient air temperature, data obtained from the PRISM Climate Group, were the most important variables impacting the final model’s accuracy.  Assuming that ambient air temperature date is available, this model allows us to predict lake photic zone temperature (Importance of this ability??).

For the final model, average ambient air temperature and the average temperature of the prior 30 days were selected.  Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.  It is likely that the three and seven day averages are not providing the model with unique information.  Typical the three and seven day averages are fairly close the day-of average.  Ambient air temperature does not change drastically over such a short time scale.  Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.  Yet the longer term 30-day average does have an impact on lake photic zone temperature.  The 30-day average provides us with this information about the temperature intensity leading up to the sample date.  Not just how far along into summer; that information would be captured in the date parameter.  But more specifically, this measures the long term thermal heating happening at a site.

In addition to the several average air temperature variables, we included land-use/land cover variables.  Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this measured).  This variables were included based on the assumption that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.  Yet neither of these variables were selected in the final model.  Even though the land-use variables were not selected that does not mean development and impervious were not impactful.  This urban-heat effect on lakes may have been adequately captured in the average ambient air temperature.  Therefore, making the land-use variables redundant.  Regardless these variables did not independently contribute to the model’s accuracy.
